---
title: "Peer Assessment 1, Reproducable Research"
output: html_document
---

**This document is for Course Project 1 of Reproducable Research course.**

## Loading & preprocessing the data

---


```{r setup}
#remove variables from environment
rm(list=ls(all=TRUE)) 
```

* Load the data  and transform it into a format suitable for your analysis


```{r}
dat = read.csv('activity.csv', header = T)
names(dat)
str(dat)
head(dat)
```

*summarize the data by day
```{r}
library(data.table)
dat_tbl = data.table(dat)
dat_tbl_summary = dat_tbl[, list(total_steps = sum(steps, na.rm = T)), 
                          by = date]
```

## What is mean total number of steps taken per day?

---


**Make The Histogram; Report on Mean and Median**


```{r}
#Creating a function
gen_hist = function(x, title){
        hist(x, 
             breaks = 20,
             main = title,
             xlab = 'Total Number of Steps', col = 'grey',
            
             cex.main = .9)
        
        #caluclate mean and median
        mean_value = round(mean(x), 1)
        median_value = round(median(x), 1)
        
        #place lines for mean and median on histogram
        abline(v=mean_value, lwd = 3, col = 'blue')
        abline(v=median_value, lwd = 3, col = 'red')
        
        #create legend
        legend('topright', lty = 1, lwd = 3, col = c("blue", "red"),
               cex = .8, 
               legend = c(paste('Mean: ', mean_value),
               paste('Median: ', median_value))
               )
}

gen_hist(dat_tbl_summary$total_steps, 'Number of Steps Taken Per Day')
```




## What is the average daily activity pattern?

----------------------------


The below plot addresses the following items:

1.  Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
2.  Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?


```{r}
#summarize dataset by interval
dat_tbl_summary_intv = dat_tbl[, list(avg_steps = mean(steps, na.rm = T)), 
                          by = interval]
#plot the time series
with(dat_tbl_summary_intv, {
        plot(interval, avg_steps, type = 'l',
             main = 'Average Steps by Time Interval',
             xlab = '5 Minute Time Interval',
             ylab = 'Average Number of Steps')
        })
#Find interval for the maximum average steps
max_steps = dat_tbl_summary_intv[which.max(avg_steps), ]

#Generate lebel
max_lab = paste('Maximum Of ', round(max_steps$avg_steps, 1), ' Steps \n On ', max_steps$interval, 'the time interval', sep = '')

#CollectcCooridinates of max interval
points(max_steps$interval,  max_steps$avg_steps, col = 'red', lwd = 3, pch = 19)

#Add Label To Annotate Maximum # Steps And Interval
legend("topright",
       legend = max_lab,
       text.col = 'red',
       bty = 'n'
       )

```

---

## Imputing missing values

* Calculate & Report The Number of Missing Values
```{r}
sum(is.na(dat$steps))
```
The steps involved in tis analysis are:
1. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

2. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
#First I will join the dataframe I created earlier that summarizes the average number of steps per interval to the original dataset
setkey(dat_tbl, interval)
setkey(dat_tbl_summary_intv, interval)


#Create function that will return the second value if the first value is NA
NA_replace = function(x,y){
        if(is.na(x)){
                
                return(y)
        }
        return(x)
}

#create new dataset that replaces NAs with average values
dat_tbl_miss = dat_tbl[dat_tbl_summary_intv]
dat_tbl_miss$new_steps = mapply(NA_replace,dat_tbl_miss$steps, dat_tbl_miss$avg_steps)

#summaryize new dataset by day
dat_tbl_summary_miss = dat_tbl_miss[, list(new_steps = sum(new_steps, na.rm = T)), 
                          by = date]
#preview new dataset
head(dat_tbl_summary_miss)
```

4.  Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

**Mean and Median Are Reported In Legend Of Histogram**

```{r}

gen_hist(dat_tbl_summary$total_steps, 'Missing Values Removed')
gen_hist(dat_tbl_summary_miss$new_steps, 'Missing Values Replaced With \n Mean For Interval')

```

**Answer To Question:**
The mean and the median are now almost the same after replacing missing values with the mean value for the relevant interval. It makes sense that the median value would now move closer to the mean. So the Median value increased after this method of missing value replacement.

## Are there differences in activity patterns between weekdays and weekends?

---
1.  Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
#Make Function To Return Either "Weekday" or "Weekend"
weekpart = function(x){
        if(x %in% c('Saturday', 'Sunday')){
                return('Weekend')
        }
        
        return('Weekday')
}

#Add Name of Week
dat_tbl_miss$dayname = weekdays(as.Date(dat_tbl_miss$date))

#Add Factor Variable To Differentiate Weekday and Weekend
dat_tbl_miss$daytype = as.factor(apply(as.matrix(dat_tbl_miss$dayname), 1, weekpart))

#Summarize Dataset: Mean grouped by interval and daytype
dat_tbl_summary_miss = dat_tbl_miss[, list(avg_steps = mean(new_steps, na.rm = T)), 
                          by = list(interval, daytype)]

#inspect dataset
str(dat_tbl_summary_miss)
```

Below is the panel plot:
```{r}
library(lattice)
xyplot(avg_steps~interval | daytype, data = dat_tbl_summary_miss,
      type = 'l',
      xlab = 'Interval',
      ylab = 'Number of Steps',
      layout = c(1,2))
```

